TrenchR: an R package for transparent environmental and ecological biophysics

### 2022-03-28 ManuscriptTe.Rmd
MS fixes

#Also estimate using a spherical approximation and a simplified version of the approximation.
H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211)

#surface area from length(m)
A= sa_from_length(l=0.0211)
```

>>>>

#Also estimate using a spherical approximation and a simplified version of the approximation.
H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211, type="Gates")

#surface area from length(m)
A= surface_area_from_length(L=0.0211)
```

### 2022-03-28 Manuscript_full.Rmd
name edit
MS fixes

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail Meyer, Yutaro Sakairi,..."
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

Lauren B. Buckley^1^, Bryan Briones Ortiz^1^, Isaac Caruso^1^, Aji
John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^

1.  Department of Biology, University of Washington, Seattle, WA
    98195-1800, USA
2.

Corresponding Author: Lauren Buckley,
[lbuckley\@uw.edu](mailto:lbuckley@uw.edu){.email}

@@

1.  Much understanding of responses to climate change and variability
    relies on the assumption that organismal body temperatures are equal
    to temporally averaged air temperatures high above the ground.
    However, most organisms live near the ground and acute exposure to
    solar radiation and thermal extremes can substantially elevate their
    body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses. The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation. We offer utility functions to aid in estimating the
    organismal and environonmental parameters needed for biophysical
    ecology.
3.  TrenchR focuses on simple and modular functions so users
    can create transparent and flexible biophysical models.
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting. To further this aim, we offer a series of educational
    modules to introduce the field of biophysical ecology.

@@

Many responses of organisms and ecosystems to climate change are
inconsistent with current predictive models [@maguire2015modeling]. Some
of these predictive shortcomings stem from omitting spatial and temporal
environmental variation and how it interacts with organismal phenotypes
(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
Many analyses assume that organismal body temperatures are equal to air
temperatures at weather station height (usually 2 meters). However, air
temperatures near the ground where most organisms reside can be
considerably warmer, and absorption of solar radiation can raise body
temperatures well above air temperatures, resulting in potential
discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014;
@kearney2009]. Applications are increasingly demonstrating the
importance of accounting for microclimate variation
[@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].
Acute thermal stress events are often more relevant to the physiology
and energy balance of organisms than daily, monthly, or annually
averaged (*Abby: changed from aggregated*) environmental conditions. Yet, daily and seasonal
environmental variation is often neglected [@dillon2016introduction;
@dillon2016life]. We aim to help remedy these shortcomings by
introducing the TrenchR package
(<https://github.com/trenchproject/TrenchR>) as a component of the
TrEnCh project (<https:www.trenchproject.com>), which builds
computational and visualization tools to Translate Environmental Change
into organismal responses.

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of these
tools has not kept pace with research on climate change responses.
Adoption of biophysical approaches is expanding with recognition of
the importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr] and the increased dissemination of computational tools. 
The TrenchR package is intended to complement
the NicheMapR package, which includes sophisticated microclimate
[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
Although the Fortran source code was recently released, the complexity
of NicheMapR functions can make it difficult to understand and modify the
source code. Many of the approaches are similar between the packages,
but TrenchR provides simple functions to aid the understanding and
transparency of biophysical approaches. The TrenchR functions are
modular and easy adapt to a variety of organisms and research
questions. Although many of the TrenchR functions are general, some
components of TrenchR are best suited for ectothermic animals.

We have limited inclusion of plant specific biophysical models since
they are the focus of several R packages [tealeaves: [@muir2019tea];
plantecophys: [@duursma2015plantecophys]]. Some energy budget
calculation components of biophysical models are also available in the
ThermImage package [@tattersall2017]. Microclimate models, with an
emphasis on describing spatial variation, are also available in the
Microclima package [@maclean2019microclima].

@@

The TrenchR package aims to promote transparency and reproducibility.
We welcome contributions and corrections from users. Our package is
built using the devtools methodology
(<https://github.com/hadley/devtools>) in Github. Issues and feature
requests can be contributed in Github
(<https://github.com/trenchproject/TrenchR/>). The package is maintained
and checked using Travis CI as a tool for continuous integration.

(*Abby: maybe add the link to the reference website? trenchproject.github.io/TrenchR/*)

@@

**Utility functions** allow estimating environmental metrics that form
the basis of microclimate and biophysical models (e.g., zenith and
declination angles), facilitate modeling diurnal temperature variation,
and estimate biologically relevant aggregate metrics such as degree days
available for development.

@@

**General biophysical functions** provide generalized models of heat
exchanges between organisms and their environment so that users can
build custom biophysical models. The functions allow implementing an
energy balance of the following components of heat exchange between
organisms and their environment: *radiative heat exchanges* of solar and
thermal radiation; *convective heat exchanges* between organisms and
their surrounding fluid (air or water) driven by fluid flow; 
*conductive heat exchanges* between organisms and the ground due to physical
contact; and *evaporative heat exchanges* associated with organisms'
evaporative water loss. Additional functions aggregate these forms of
heat exchange into energy balances and to use the energy balances to
predict body temperatures.

@@

We introduce the functions in several vignettes. A good place to start
is the *Allometry and conversions* vignette, which provides tools for
preparing data, such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* vignette
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and wind speed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models.

@@

## Implementation
(*Abby: I might title this "Example Application" or "Demonstration of Use" or other. I would assume implementation to mean how you implemented the functions overall*)

@@

We will use the energy budget to estimate body temperature, $T_b$, which
can depart dramatically from the air temperatures due to energy exchange
with the environment. Heat is gained from absorbed solar and thermal
radiation and from metabolic reactions. Heat is lost through the
organism's emission of radiation and the evaporation of water. The
organism exchanges heat with the surrounding air or water via convection
and with the ground via conduction. The balance of these heat exchanges
(omitting metabolism and evaporation, which are often negligible for
ectotherms) can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement].
(*Abby: I'm confused at where the distinction is made between Tb and Te, for example if the energy exchanges are part of Tb or Te or both.*)

Let us assume the lizard is in an unshaded location where a weather
station at standard height (2 meters) reports the daily air temperature
varies from a minimum of 10°C to a maximum of 25°C and the wind speed
averages 1 m/s. The soil surface temperature varies from a minimum of
15°C to a maximum of 30°C. We assume that atmospheric transmissivity
$\tau = 0.7$ and albedo $\rho = 0.6$

@@

We first prepare the environmental data for analysis. We estimate 
hourly air and soil temperatures and radiation. We start by
estimating the day of year and the timing of sunrise and sunset:

@@

Although measured solar radiation is preferable if available, we can
estimate hourly solar radiation by discounting incoming solar radiation
as it moves through the atmosphere as follows:

@@

We use microclimate models to scale air temperature and wind speed from
weather station height (2 meters) to lizard height (0.02 meters). We assume a
surface roughness of $z_0 = 0.2$ (*Abby: units?*), which corresponds to bare sand and
determines the turbulence of airflow. We implement a neutral air
temperature profile, an air temperature
profile forced by wind speed, and a neutral wind speed profile.

@@

# Estimate surface area and sihouette area
A = sa_from_mass(mass, "lizard") # surface area (m^2)
psa = sapply(psi_deg, prop_silhouette_area, taxa = "lizard", posture = "prostrate") # Projected silhouette area (m^2)

@@

We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$
(W). We will estimate the lizard's heat transfer coefficient, $H_L$
($Wm^{-2}K^{-1}$) using an empirical relationship for lizards
(`heat_transfer_coefficient()`), but also illustrate a function
estimating $H_L$ using a spherical approximation
(`heat_transfer_coefficient_approximation()`) and a simplified
approximation (`heat_transfer_coefficient_simple()`).

@@

# Estimate the heat transfer coefficient using a simplified version of the approximation
H_L3= heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
```
*Abby: Added type=Gates above to fix error in H_L3, change if Gates is not the equation wanted. Also, maybe give a reason why one might want to use a spherical or simplified version*

@@

We additionally estimate $T_e$ using a specialized function for lizards
[@buckley2008link]:

@@

## Acknowledgements

We thank Brian Helmuth and Eric Riddell for providing and checking
functions. We thank Ray Huey, Joel Kingsolver, Luke Miller, Warren
Porter, and Art Woods for assistance and input. We particularly thank
Michael Kearney for extensive input on TrenchR and its alignment with
NicheMapR and for contributing two educational modules.

[Check with Eric]

@@

LB conceived and designed the package and lead all aspect of the
packages creation and dissemination; BBO, IC, AJ, OL, AM, YS wrote and
edited functions and associated package resources; AJ configured the
package; AM, BBO, YS, and LB converted and edited the educational
modules. All authors contributed critically to the drafts and gave final
approval for publication.

>>>>

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail V. Meyer, Eric A. Riddell, and Yutaro Sakairi"
output:
  html_document: default
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
csl: methods-in-ecology-and-evolution.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

Lauren B. Buckley^1^, Bryan A. Briones Ortiz^1^, Isaac Caruso^1^, Aji
John^1^, Ofir Levy^2^, Abigail V. Meyer^1^, Eric A. Riddell^3^, Yutaro Sakairi^1^

1.  Department of Biology, University of Washington, Seattle, WA
    98195-1800, USA
2.  School of Zoology, Tel Aviv University, Tel Aviv, Israel
3.  Department of Ecology, Evolution, and Organismal Biology, Iowa State University, 
Ames, IA 50011-4009, USA
Corresponding Author: Lauren Buckley,
[lbuckley\@uw.edu](mailto:lbuckley@uw.edu){.email}

@@

1.  Much understanding of responses to climate change and variability
    relies on the assumption that organismal body temperatures are equal
    to temporally averaged air temperatures high above the ground.
    However, most organisms live near the ground and acute exposure to
    solar and thermal radiation and thermal extremes can substantially elevate their
    body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses. The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation. We offer utility functions to aid in estimating the
    organismal and environonmental parameters needed for biophysical
    ecology.
3.  TrenchR focuses on simple and modular functions so users
    can create transparent and flexible models for biophysical applications.
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting. We further this aim through a series of educational
    modules to introduce the field of biophysical ecology.

@@

Responses of organisms and ecosystems to climate change are heterogeneous and thus
inconsistent with current predictive models [@maguire2015modeling]. Some
of these predictive shortcomings stem from omitting spatial and temporal
environmental variation and how it interacts with organismal phenotypes
(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
Many analyses assume that organismal body temperatures are equal to air
temperatures at weather station height (usually 2 meters). However, air
temperatures near the ground where most organisms reside can be
considerably warmer, and absorption of solar and thermal radiation can raise body
temperatures well above air temperatures, resulting in potential
discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014;
@kearney2009]. Applications are increasingly demonstrating the
importance of accounting for microclimate variation
[@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].
Acute thermal stress events are often more relevant to the physiology
and energy balance of organisms than daily, monthly, or annually
averaged environmental conditions. Yet, daily and seasonal
environmental variation is often neglected [@dillon2016introduction;
@dillon2016life]. We aim to help remedy these shortcomings by
introducing the TrenchR package
(<https://github.com/trenchproject/TrenchR>) as a component of the
TrEnCh project (<https:www.trenchproject.com>), which builds
computational and visualization tools to Translate Environmental Change
into organismal responses.

The mathematical tools for leveraging environmental data to predict
organismal conditions have long lingered in books
[@campbell2000introduction; @gates1980biophysical] and articles
[@porter1969thermodynamic; @porter1983biophysical], but adoption of these
tools has not kept pace with research on climate change responses.
Adoption of biophysical approaches is expanding with recognition of
the importance of spatial and temporal environmental variation to
determining climate change responses [@buckley2018leveraging;
@kearney2017nichemapr] and the increased dissemination of computational tools.

The TrenchR package is intended to complement
the NicheMapR package, which includes sophisticated microclimate
[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
Although the Fortran source code was recently released, the complexity
of NicheMapR functions can make it difficult to understand and modify the
source code. Many of the approaches are similar between the packages,
but TrenchR provides simple functions to aid the understanding and
transparency of biophysical approaches. The TrenchR functions are
modular and easy adapt to a variety of organisms and research
questions. Although many of the TrenchR functions are general, some
components of TrenchR are best suited for ectothermic animals. We have limited inclusion of plant specific biophysical models since
they are the focus of several R packages [tealeaves: [@muir2019tea];
plantecophys: [@duursma2015plantecophys]]. Some energy budget
calculation components of biophysical models are also available in the
ThermImage package [@tattersall2017]. Microclimate models, with an
emphasis on describing spatial variation, are also available in the
Microclima package [@maclean2019microclima].

@@

The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to promote transparency and reproducibility.
We welcome contributions and corrections from users. Our package is
built using the devtools methodology
(<https://github.com/hadley/devtools>) in Github. Issues and feature
requests can be contributed in Github
(<https://github.com/trenchproject/TrenchR/>).

@@

**Utility functions** allow estimating environmental metrics that form
the basis of microclimate and biophysical models (e.g., zenith and
declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation,
and estimate biologically relevant aggregate metrics such as degree days
available for development.

@@

**General biophysical functions** provide generalized models of heat
exchanges between organisms and their environment so that users can
build custom biophysical models. The functions allow implementing an
energy balance including the following components of heat exchange between
organisms and their environment: *radiative heat exchanges* of solar and
thermal radiation; *convective heat exchanges* between organisms and
their surrounding fluid (air or water) driven by fluid flow; 
*conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical
contact; and *evaporative heat exchanges* associated with organisms'
evaporative water loss. Additional functions aggregate these forms of
heat exchange into energy balances and to use the energy balances to
predict body temperatures.

@@

We introduce the functions in several vignettes. A good place to start
is the *Allometry and conversions* vignette, which provides tools for
preparing data, such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* vignette
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and wind speed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models. We additionally offer a *List of symbols* used in equations.

@@

## Example Application

@@

We will use the energy budget to estimate body temperature, $T_b$, which
can depart dramatically from the air temperatures due to energy exchange
with the environment. Heat is gained from absorbed solar and thermal
radiation and from metabolic reactions. Heat is lost through the
organism's emission of radiation and the evaporation of water. The
organism exchanges heat with the surrounding air or water via convection
and with the ground via conduction. The balance of these heat exchanges
(omitting metabolism and evaporation, which are often negligible for
ectotherms) can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement]. $T_e$ is an estimate of $T_b$ and the package functions refer to $T_b$ for simplicity.

Let us assume the lizard is in an unshaded location where a weather
station at standard height (2 meters) reports that the daily air temperature
varies from a minimum of 10°C to a maximum of 25°C and the wind speed
averages 1 m/s. The soil surface temperature varies from a minimum of
15°C to a maximum of 30°C. We assume that atmospheric transmissivity
$\tau = 0.7$ and albedo $\rho = 0.6$

@@

We first prepare the environmental data for analysis. We will estimate 
hourly air and soil temperatures and radiation usinf a function describing diurnal temperature variation. We start by
estimating the day of year and the timing of sunrise and sunset:

@@

Although measured solar radiation is preferable if available, we can
estimate hourly solar radiation by discounting incoming solar radiation
as it moves through the atmosphere as follows. The function partitions radiation into direct, diffuse, and reflected components:

@@

We use microclimate models to scale air temperature and wind speed from
weather station height (2m) to lizard height (0.02m). We assume a
surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and
determines the turbulence of airflow. We implement a neutral air
temperature profile, an air temperature
profile forced by wind speed, and a neutral wind speed profile.

@@

# Estimate surface area and sihouette area
A = surface_area_from_mass(mass, "lizard") # surface area (m^2)
psa = sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate") # Projected silhouette area (m^2)

@@

We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$
(W). We will estimate the lizard's heat transfer coefficient, $H_L$
($Wm^{-2}K^{-1}$) using an empirical relationship for lizards
(`heat_transfer_coefficient()`). We also illustrate a function
estimating $H_L$ using a spherical approximation
(`heat_transfer_coefficient_approximation()`) and a simplified
approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.

@@

# Estimate the heat transfer coefficient using a simplified version of the approximation
H_L3= heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
```

@@

We additionally estimate $T_b$ using a specialized function for lizards
[@buckley2008link]:

@@

## Acknowledgements
We thank Ray Huey, Joel Kingsolver, Luke Miller, Warren
Porter, and Art Woods for assistance and input and Brian Helmuth for providing functions. We particularly thank
Michael Kearney for extensive input on TrenchR and its alignment with
NicheMapR and for contributing two educational modules.

@@

LB conceived and designed the package and lead all aspect of the
packages creation and dissemination; BBO, IC, AJ, OL, AM, ER, YS wrote and
edited functions and associated package resources; AJ configured the
package; AM, BBO, YS, and LB converted and edited the educational
modules. All authors contributed critically to the drafts and gave final
approval for publication.
